Impact of grids and dycores in CESM2.2 on the meteorology and climate of the Arctic

### 2020-11-04 arcticgrids.tex
ok

General Circulation Models (GCMs) are a powerful tool for understanding the meteorology and climate of the poles, which are among the most sensitive regions on Earth to global and environmental change. Despite their importance, not all GCMs treat the poles the same way numerically as they do other regions of the globe, owing to the \textit{pole-problem} \cite{W2007JMSJ}. The pole-problem refers to instability arising from the convergence of meridians into a polar singularity on latitude-longitude grids. In the 1970's this issue was largely defeated through wide-spread adaption of the spectral transform methods in GCMs, which transforms grid point fields into an isotropic representation in wave space. But as computing power has increased, local numerical methods have become more desirable for their ability to run efficiently on massively parallel systems. The pole-problem has thus re-emerged in contemporary climate models using latitude-longitude grids, and some combination of reduced grids and polar filters are necessary to subdue this instability \cite{JW2010LNCSE}. An alternative approach is to use unstructured grids. Unstructured grids permit quasi-uniform grid spacing globally, thereby eliminating the pole-problem. They allow for more flexible grid-structures than latitude-longitude grids, and can support regional grid refinement that may be used to capture higher resolution in the polar regions. Regional refinement of polar regions may be desirable as latitude-longitude grids, by virtue of the convergence of meridians, have higher horizontal resolution in polar regions compared to quasi-uniform grids containing the same degrees of freedom.

@@

The Arctic region has a unique geography. It consists of a polar ocean encircled by the northern coasts of the northern hemisphere continents and the large island of Greenland, and capped by sea ice for half the year. Greenland is covered in a kilometers thick ice sheet, and receives a substantial portion of its snowfall from synoptic-scale storms traversing up the eastern sea-board of North America, and steered in the direction of the Icelandic low. Katabatic winds accelerate down the steep slopes of the Greenland Ice Sheet (GrIS) in episodic bursts that can clear out sea ice from its fjords and surrounding coastal region. Large volumes of fresh water are drained into the ocean through river outlets in Canada and Russia, ice discharge and melt water runoff from the periphery of GrIS. Polar Lows form during the cold season and their gale-speed winds can kick up large waves and pose a hazard to mariners' and coastal communities. These Arctic processes have proven challenging to represent in GCMs, primarily due to the fine-scales involved \cite{BETAL2001MWR,SG2017MWR,VETAL2018TC}. As such, the Arctic serves as a test-bed for comparing the simulated meteorology and climate in the SE and FV dycores.

>>>>

General Circulation Models (GCMs) are a powerful tool for understanding the meteorology and climate of the poles, which are among the most sensitive regions on Earth to global and environmental change. Despite their importance, numerically, not all GCMs treat the polar regions the same way as they do elsewhere in the domain due to the \textit{pole-problem} \cite{W2007JMSJ}. The wide variety of grids structures and numerical techniques used to represent polar regions in conventional GCMs are evaluated in this study for their impacts on the meteorology and climate of the Arctic.

The pole-problem refers to instability arising from the convergence of meridians into a polar singularity on latitude-longitude grids. In the 1970's this issue was largely defeated through wide-spread adaption of the spectral transform methods in GCMs, which transforms grid point fields into an isotropic representation in wave space. But as computing power has increased, local numerical methods have become more desirable for their ability to run efficiently on massively parallel systems. The pole-problem has thus re-emerged in contemporary climate models using latitude-longitude grids, and some combination of reduced grids and polar filters are necessary to subdue this instability \cite{JW2010LNCSE}.

An alternative approach is to use unstructured grids. Unstructured grids permit quasi-uniform grid spacing globally, thereby eliminating the pole-problem. They allow for more flexible grid-structures than latitude-longitude grids, and can support regional grid refinement that may be used to capture higher resolution in the polar regions. Regional refinement of polar regions may be desirable as latitude-longitude grids, by virtue of the convergence of meridians, have higher horizontal resolution in polar regions compared to quasi-uniform grids containing the same degrees of freedom.

The meteorology and climate of the Arctic is charcterized by a range of processes and scales that are difficult to represent in GCMs \cite{BETAL2001MWR,SG2017MWR,VETAL2018TC}. The Arctic geography consists of a polar ocean surrounded by the northern coasts of the northern hemisphere continents and the large island of Greenland. Greenland is covered in a kilometers thick ice sheet, and receives a substantial portion of its snowfall from synoptic-scale storms traversing up the eastern sea-board of North America that rain out over its steep margins. Intense cold season Polar Lows traverse the Arctic's melange of sea-ice and open ocean, inducing gale-speed winds. Large volumes of fresh water drain into the ocean through river outlets in Canada and Russia, iceberg discharge and melt water runoff emanating from the margins of GrIS. As such, the Arctic serves as a test-bed for comparing the simulated meteorology and climate to different numerical treatments of polar regions.

@@
